Speculation in foreign exchange: noise or information

### 2014-11-13 Knitr/Paper.tex
Updated the noise document with some deletions and additionas of new reading.  Needs another look and review

\documentclass[12pt, a4paper, oneside]{article} % Paper size, default font size and one-sided paper
%\usepackage[dcucite]{harvard}
\usepackage{tikz}
\usetikzlibrary{shapes, shadows, arrows}
\usepackage{rotating}
\usepackage{amsmath}
%\usepackage{setspace}
\usepackage{pdflscape}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage[comma, sort&compress]{natbib}% Use the natbib reference package - read up on this to edit the reference style; if you want text (e.g. Smith et al., 2012) for the in-text references (instead of numbers), remove 'numbers' 
\usepackage{graphicx}
\graphicspath{{./Figures/}} % Specifies the directory where pictures are stored
%\bibliographystyle{plainnat}
\usepackage{listings}
\bibliographystyle{agsm}
\usepackage[colorlinks = true, citecolor = blue, linkcolor = blue]{hyperref}
%\hypersetup{urlcolor=blue, colorlinks=true} % Colors hyperlinks in blue - change to black if annoying
\begin{document}
%\renewcommand[\harvardurl]{URL: \url}
\title{Speculation in foreign exchange: noise or information}
\author{Rob Hayward\footnote{University of Brighton Business School, Lewes Road, Brighton, BN2 4AT; Telephone 01273 642586.  rh49@brighton.ac.uk.  We are very grateful for the comments and suggestions made.  In particular, the EACES conference in Budapest between September 5th and 6th 2014, Andros Steinhoffer and.....}} 
\date{\today}
\maketitle
\begin{abstract}
An event study of extreme speculation is used to test the informational content of speculation.  Speculative extremes are ranked by weight of activity and intensity of sentiment.  In either case, the extreme says little about the future, suggesting that speculation is more than just noise and that there is some fundamental informational content. A unique dataset of risk-reversal skew on option prices is used to measure the weight of speculative sentiment.   
\end{abstract}

@@

There is a fundamental schism in the literature between those that view speculation as uninformed noise activity that can run in waves when affected by common decision-making bias or social influences, and those that view speculation as activity informed by fundamental knowledge that will stabilise markets.  In the former case, there is a risk that speculative activity will drive prices away from fundamental value as price changes and sentiment feed off each other.  At its most extreme, speculation in this interpretation is likely to turn into a bubble that may burst in a cascade where the previous run-up in price and sentiment is swiftly reversed.  However, the alternative view of speculation as an activity that is driven by fundamental knowledge and information would suggest that speculation acts to limit overshooting and divergence from fundamental value, thereby reducing volatility.  This second view says that it is the absence of speculators and the liquidity that they provide that encourages excess price movement and volatility.  The aim of this chapter is to use the foreign exchange market to assess the relationship between intensity or weight of speculation and the movement of prices.  If speculators are uninformed, their most extreme sentiment and their greatest weight in market activity should tend to cause overshooting and presage bubble-bursting, reversals and a crash; if speculators are informed, activity will drive prices towards fundamental value and equilibrium where volatility and crash risk should be reduced.

\subsection{Fundamental value}
Fundamental or intrinsic value is a difficult concept to apply to foreign exchange.     There is a weight of empirical evidence that standard economic models are not very effective in explaining the behaviour of exchange rates.  Meese and Rogoff found that in out-of-sample forecasts, a random walk performed as well as the four major economic models of foreign exchange behaviour that were employed \citep{Meese1983Empirical}.  These findings have been repeated in a number of similar, subsequent studies  \citep{Frankel1984tests},  \citep{Frankel1990Exchange}, \citep{Lyons1995Microstructure} and \citep{Sarno2006}.  In the previous chapter, the fundamental value of the exchange rate was assessed to be PPP.  However, even there the real exchange rate was not stable (see Figure \ref{fig:timeseries} and view the variable RTWI to see the evolution of the US real trade weighted index over the period 1985 to 2010).  Therefore, while the law of one price and purchasing power parity generally determines exchange rates in the long-run for currencies of countries at a similar level of economic development, in the short to medium-term there is still no robust economic model that can consistently explain exchange rate movements, let alone provide a forecast.

Noise-trading is the label attached to that trading or investment where decisions seem to have been made with no reference to fundamental information; activity is like a random noise.  This chapter provides an investigation into the effect of speculation in the foreign exchange market.  It seeks to assess the relationship between sentiment, speculation, information and exchange rate prices.  The DeLong-Summers model of noise trading is calibrated against daily and weekly returns for major foreign exchange rates \citep{Delong1990noise}.  Little support for the hypothesis that extreme levels of sentiment or extreme levels of speculative activity convey information about the future direction of prices is found.  This is something of a contrast to the evidence that has been found in some other markets.  See  \citep{DeBondtOver}, \citep{Chopra1992} and \citep{SteinOptions} amongst others (discussed more fully in Section \ref{secref:empiric} below).   The results of the experiment carried out here suggest that it is not sufficient just to measure the intensity of speculative sentiment or the weight of speculators to determine the divergence from intrinsic value and that the relationship between speculation and price is rather complex.  There is evidence here against the idea of speculation as purely noise.  Rather, it appears that speculation is part of the price discovery process.

\section{Literature review}  
In many ways this is a study about how efficient the market  is in processing information, specifically whether the presence of speculators improves or reduces the efficiency with which information is processed.  As such, there must first be some discussion of information and ideas about the efficiency with which it is incorporated into asset prices.

Bachelier is responsible for the first structured  study of informational efficiency in his 1900 doctoral analysis of the behaviour of French government bond, future and option prices \citep{Bachelier1900}.  His is apparently the first analysis that suggested that the evolution of securities' prices was a \emph{Martingale}, meaning that past events do not help to predict the future value, that the expected future value is equal to the current value and that the process is a \emph{fair game} for which the probability of winning or losing is equal.  A \emph{random walk} is a slightly stronger construct as each step is an independent and identically distributed random variable.  However, a martingale may contain past information about the higher moments.   So there may be, for example, volatility clustering. Brownian motion is named after the Scottish botanist Robert Brown and refers to the random movement of particles suspended in a liquid or gas.  See \citep{LeRoyMartingales} for an overview of market efficiency, random walks and Martingale processes.  Bachelier studied a set of bond prices and found that the potential returns of these $19^{th}$ century French government bonds could be represented by a probability distribution where the mean was zero once the gravitation towards parity at maturity and seasonal cycle of `dirty prices' with accrued interest had been accounted for.

\begin{quotation}
`` The determination of these fluctuations depends on an infinite number of factors; it is, therefore impossible to aspire to mathematical prediction of it.  Contradictory opinions concerning these changes diverge so much that at the same instant buyers believe in a price increase and sellers in a price decrease."
\end{quotation}

\citep[p. 17]{Bachelier1900}

Bachelier's ideas received little immediate attention and remained dormant until renewed interest in the process describing securities prices developed in the 1950s.  However, others were tentatively pursuing similar investigations.  In the 1930s Working assessed the empirical evidence for random behaviour of stock and commodity prices \citep{Working1934} and U.K. statistician Kendall studied the performance of weekly price data from the Dow Jones \citep{Kendall1953}.  Among those taking an interest in the subject were Roberts, who scorned the \emph{patterns} that were identified by professional investors while making the case for a focus on the change in price of securities rather than the price itself.  What we now call \emph{returns} were, argued Roberts, much more random than price levels.  He calibrated a model of random shocks to the price change, using the RAND Corporation's random number text \citep{RAND} and created graphs that were remarkably close to those produced by real stocks.  Nonetheless, Roberts warned his research would not prevent people from using past data to try to make investment decisions as ``chance behaviour itself provides 'patterns' that invite spurious interpretations'' \citep[p. 2]{Roberts1959}.

Towards the end of the decade, Osborne made the direct comparison between the price movement of stocks and the activity of small particles through the analysis of prices of a selection of stocks from the American and New York stock exchanges.  The data were collected from \emph{The Wall Street Journal} by a research company called F.W. Stephens for the period from 1924 to 1956.  The log of the price ratio for a range of different time periods and their distribution was compared to what would have been expected from something evolving with Brownian motion.  By plotting the ratio over time, Osborne showed that changes in the log of the price ratio was close to being normally distributed.  Osborne argued that the securities prices jumped around to find the point where buyers and sellers were balanced and overall opinion about the direction of prices was close to zero. Therefore, the evolution of prices was a like a sequence of independent random variables that were the consequence of transactions taking place.  The distribution function for this sequence of random variables was, according to Osborne's study, normal with a zero mean and a variance that increased steadily with time.  Osborne concluded that the random nature of price changes would ensure that buyers and sellers could only make a standard return on average  \citep{Osborne1959}.

Samuelson provided a theoretical foundation for the idea that securities prices would follow a random walk.  Starting with a series of simple axioms, he deduced that competitive markets would drive securities prices towards the expected value, where the expected value is defined as the projected mean of the process underlying the evolution of the price.  This introduced the idea of a \emph{Martingales} or the notion that the expected value of a security or financial asset would be equal to its current value.  If that is not the case, rational investors in a competitive market will take advantage of the deviation by buying when it is below the expected value and selling when it is above and, as a result of this activity, would drive the price towards the expected level and, in the process, remove the opportunity to make profits.   Using the backward iteration of expectations, the theory states that a stock price should equal the expected present value of future dividends. This work provided the underpinning and stimulus for another wave of empirical examinations but rests, of course, on the assumptions made about the competitive nature of markets and the ability of investors to form expectations about the future \citep{SamuelsonStockPrices}.

Amongst these empirical investigations, Cootner tested Bachelier's ideas on a range of financial markets, most notably the forward market for grain \citep{CootnerSpeculators} and  Fama found that the returns of stocks in the Dow Jones Industrial Average were very close to the benchmark of a random walk \citep{FamaStock}.  At this point the term \emph{Efficient Market Hypothesis} (EMH) became more widely used.  Jensen described market efficiency as

\begin{quotation}
``A market is efficient with respect to information set $\Omega_t$ if it is impossible to make economic profits by trading on the basis of information set $\Omega_t$".  
\end{quotation}
\citep[p. 95]{JensenEfficiency}.

More formally,

\begin{equation}
E[R_{t+1}, Q_{t+1} | \Omega_t] = 0
\end{equation}

where $E$ is the expectations operator, $Q_{t+1}$ is the risk premium at time $t+1$, $R_{t+1}$ is the return at time $t +1$ and $\Omega_t$ is the information set at time $t$ \citep[p. 95]{JensenEfficiency}.  Jensen also said "I believe that there is no other proposition in economics which has more solid empirical evidence supporting it than the Efficient Market Hypothesis" \citep[p. 93]{JensenEfficiency}.   Jensen was, however, identifying a number of inconsistencies and anomalies that had been uncovered and was anticipating a revolution that would overthrow the theory.  He was anticipating the paradigm shift that was to follow.

%this paragraph goes somewhere else. 
%Among these were positive returns following an earnings release \citep{Ball1978} and \citep{Watts1978};  a successful trading rule for buying closed-end funds trading at a discount to asset value \citep{Thompson1978}; discrepancies between stock and option prices on the NYSE \citep{Galai1978}; and, information in option prices \citep{ChirasManaster}.

\subsection{Failures of market efficiency}
Questions about the EMH had emerged gradually with doubts about some of the assumptions that Samuelson had used in the construction of his theory and with a rising number of indications that financial asset prices did not evolve in a totally random fashion.  There were three main strands to the question that arose:   empirical evidence that the price of securities did not follow a random walk made up of a series of independent and identically distributed random variables;  questions about the effects of institutional structures of financial markets on price;  the criticism of VNM expected utility theory as a valid model of investment decision-making.

\subsubsection{Empirical evidence}
The empirical evidence that has been gathered about the nature of the returns to securities reveals two primary issues.  The first is that serial correlation or trends are evident and the second is the leptokurtic and skewed distribution of these returns.  The former is the most damaging to the idea that financial markets can be informationally efficient as it suggests, contrary to the implications of the EMH, that future price changes could be partially foreseen by the movement of past price changes.\footnote{This of course depends on the assumption of \emph{risk neutrality} and the notion that investor decisions will be determined by the first two moments of the distribution of expected returns.  If the return distribution is skewed or fat tailed or there is serial correlation in the conditional variance of the returns, there may be unexplored opportunities to reduce risk or to increase risk-adjusted returns. The third and fourth moments of the distribution of returns can be extremely important in assessing the level of risk.}  The evidence seems to support the notion that prices can evolve in trends, and that there are times when positive price developments are more likely to be followed by further positive developments and that there are other times when positive price developments are more likely to be followed by a negative outcome.

There are a number of studies that find evidence that the reaction of stocks to information is less than instantaneous.  Ball and Brown seems to be the first study that identifies the upward drift in cumulative abnormal returns after some positive earnings news and downward drift in cumulative abnormal earnings after negative earnings news \citep{BallBrown}.  Bernard provides a summary of the evidence that stocks react slowly to earnings announcements.  He cites twenty examples of ``post earnings announcement drift'' \citep[p. 303]{BernardDrift}.  Part of this under-reaction comes from simply failing to identify what current earnings mean for subsequent quarters.    Bernard argues that the evidence is firmly on the side of earning reactions being too small and taking on average six months to be fully digested \citep[p. 305]{BernardDrift}.

@@

\subsubsection{Normal distribution}
The indications that the distribution of financial security returns are likely to be leptokurtic and skewed would not necessarily undermine the EMH as long as the distributions were independent.  However, this evidence that the distribution of returns is not normal, reinforces the view that standard analysis of investment based on the first two moments of the return distribution may not be adequate to understand risk.   This is something that will be more fully investigated in Chapter \ref{Chapter4} with a study of the carry trade.  In addition, if there is a pattern to the evolution of the higher moments over time, which appears to be the case, this can also suggest that there may be informational inefficiencies that can be exploited to make abnormal returns.

For example, volatility clustering means that a period when the returns to financial assets exhibit a high variance is much more likely to be followed by another period with high volatility than one where the volatility returns to some average level.  The creation of autoregressive conditional hetroscedasiticity (ARCH) and general autoregressive  conditional hetroscedasticity (GARCH) models have been designed to understand this process and provide greater precision about the effect that a shock to volatility will have on the system.  See \citep{Engle1982} for the original model, \citep{Bollerslev1986} for the generalised version, \citep{Engle1987} for an application to time-varying risk premia and \citep[pp. 657 - 665]{Hamilton} for a fuller overview of the subject.

@@

In 1959 Osborne found that returns were nearly normally distributed.  However, the quantile plots that he produced showed that there were slightly more extreme readings than would be expected \citep{Osborne1959}.  This was a finding that was repeated by a number of other studies.  With computer power lacking and measurement imprecise, these finding of a small deviation from normality were initially ignored.  Paul Cootner had already found that extreme returns were more likely to be recorded at the daily frequency than at the monthly level.  This, he suggested was related to the increased ability of speculators to take action at the lower frequency to limit extreme price movement \citep{Cootner1962}. However, the mathematician Benoit Mandelbrot rejected this idea, suggesting that the discrepancy between the evidence from the different frequencies was due to sample size (there being much more high-frequency data).  It was not what he had found with cotton prices.  Mandelbrot had made his own investigation of cotton prices and suggested that serial correlation in the returns was a significant problem and that an alternative distribution would be more appropriate for the returns as there were a number of extreme readings \citep{Mandelbrot1963} and \citep{Mandelbrot1971}.

@@

% Mandelbrot uses what he calls a \emph{stable Paretian} distribution to model returns.  This is a version of Levy's 1925 model and is the same as that used in chapter 4 by ref.  GED or something?  It is a general, stable distribution function with the parameters alpha, beta, x ? representing the tails, skewness, location and wideth.  A beta below 2 suggests flat tails in the distrubution. This is the distrubiton that Mandelbrot and Fama proposed. 
\subsubsection{Decision-making} 
As discussed in Section \ref{sec:prob} there are alternative views of the decision-making process.  Knight, Shackle, Keynes and others had suggested that that not all decisions could be subjected to a quantifiable, probabilisitic approach.  In addition, Simon updated the idea of \emph{economic man} with \emph{global rationality} to something that was consistent with the actual quality of information and computational capacities of economic agents.  Simon questioned the model of the \emph{economic man} with stable preferences, clear and voluminous information and computational skill.  Simon introduced the term \emph{bounded rationality} to express the limits to VNM rational-decision-making: there is only a certain amount of information that can be processed and probably only a certain amount of information that is available  \citep[p. 99]{Simon1955Behavioural}.    Simon wanted to use psychological evidence to support his ideas, but found it lacking at that time.  However, as more experimental work on the psychology of decision-making was carried out, empirical support was provided for many of the themes that Knight, Shackle, Keynes and Simon had already proposed.

%Of course, even those involved in the original construction of the EMH were aware that this was a model rather than a description of the decision-making process.  For example, back in 1959 Osborne stated that investment decisions would be made "as if" calculating the expected change in the log ratio of prices and

%\begin{quotation}
%"From what has been said about the anatomy of logical decisions (they need not be consciously logical, but this is the supposition as to how they are reached), we must suppose that for the buyer, his estimate of the expectation value of the change in value $(\Delta log_e P)$ is positive while the sellers estimate of the same quantity is negative"
%\end{quotation}
%\citep[p. 153]{Osborne1959}

%What is important is whether this model provides outcomes that are consistent with what is evident.  Again, it may be the case that in the long-term the standard model of rational decision-making provides a reasonable description of what happens in aggregate but there is a risk that at the individual level and on the path to the equilibrium, the process of groping around with limited information and limited ability to assess possible outcomes can create outcomes that diverge from that of the standard model.

%This needs to find a place in the introduction.   
%Herbert Simon introduced the term 'bounded rationality' to express the limits to rational-decision-making: there is only a certain amount of information that can be processed and probably only a certain amount of information that is available.  Simon was focused more on the decision-making process itself than the aggregate outcome of decisions that had been made.  He questioned the model of the "economic man" with stable preferences, clear and voluminous information and computational skill.   One of the key ideas that he developed was that of 'satisficing' rather than maximising utility.  It means that decisions are taken to achieve at least a particular outcome. In the field of investment, it may mean that decisions are taken to ensure that a minimum level of return is achieved rather than the maximum. This can reduce the amount of options that have to be assessed \citep{Simon1955Behavioural}.

%As we have seen, Knight, Shackle and Keynes %ref for introduction
%were particularly sceptical about rational decision-making.  Keynes argued that given the uncertainty that is attached to long-term investment projects, it is likely that investors will tend to use a short-cut to decision making that would be based on the assumption that the valuation or price will remain unchanged unless there is some strong reason to believe that it should be elsewhere.  Keynes' own statement of the EMH is 
%\begin{quotation}
%We are assuming, in effect, that the existing market valuation, however, arrived at, is uniquely correct in relation to our existing knowledge of the facts which will influence the yield of the investment, and that it will only change in proportion to changes in this knowledge. 
%\end{quotation}
%\citep[p. 152]{Keynes1936}.

%However, Keynes questioned this notion, arguing that the existing market valuation was more likely to be influenced by a variety of ephemeral, short-term fluctuations in sentiment and popular opinion with market valuations being based more on ignorance and subject to sudden fluctuations in the mass interpretation of the available information.  As a result, he felt that "the market would be subject to waves of optimistic and pessimistic sentiment" \citep[p. 33]{Keynes1936}.  One important aspect of this is the influence of public opinion and the importance of social pressures.  Keynes argues that the investment process is like a competition where the winner is the one that can guess what everyone else will choose.\footnote{The third degree "where we devote our intelligence to anticipating what average opinion expects average opinion to be" \citep[p. 156]{Keynes1936} has been achieved with 'sniffer algorithms' which try to identify the patterns used by other machines so that they can get a head start on opponents strategies.  See, for example, More or Less, 11 August 2011, BBC 4.}

\subsubsection{Behavioural finance}
In 2002 Daniel Kahneman shared the Nobel prize in economics for the work that he did with Amos Tversky. The Nobel committee cited the application of experiments and surveys to question the assumptions of economic rationality and the model of VNM expected utility and to build a more empirically-based theory of how decisions are made as a justification for the award \citep{Kahneman1979Prospect} and \citep{KahnemanThinking}.  Kahneman and Tversky present \emph{Prospect Theory} which is an alternative to the VNM expected utility theory.  The theory was developed from the results of a large number of experiments that Kahneman and Tverskey carried out to ask people what would be their decision given a choice of alternatives.  Kahneman and Tverskey then compared their results to those that would be expected from a VNM expected utility model.

Kahneman and Tversky developed a \emph{decision-weighting function} $\pi = \pi(p)$ which had a similar effect as the adjustments that Keynes makes to standard probability estimates using $q$ to cover risk aversion and $w$ to cover uncertainty. The essential difference between Prospect Theory and the VNM expected utility benchmark are that people underweigh outcomes that are probable compared to certainties, which contributes to relatively high risk-aversion in choices involving sure gains and risk-taking in choices involving sure losses \citep[p. 263]{Kahneman1979Prospect}.  In practical terms, for investors, this encourages the phenomenon of selling winners early to lock in gains and holding on to losers for too long, gambling in the face of adverse consequences.  In addition, the people making decisions in the experiments conducted by Kahneman and Tversky are concerned about changes in wealth rather than the level of wealth itself.  According to Kahneman and Tverskey, this explains why people are prepared to purchase insurance that will cover losses that are a relatively small proportion of their wealth, losses that they could presumably self-insure without the administration costs of a formal insurance company \citep[p. 286]{Kahneman1979Prospect}.

%can we cover what Shackle, Knight, Keynes and Simon said here....
More relevant for this study are the \emph{heuristics} or short-cuts to decision-making that have been identified by Kahneman, Tversky and other psychologists.   A range of investigations reveal that people deal with the problem of limited information and limited cognitive processing power by using these heuristics  that will economise on time and effort.  The heuristics allow decision to be made quickly with a limited amount of information.  However, they may also create some systematic biases in decision-making that can ensure that, in some situations, investors will make systematic mistakes in their evaluations.  If the mistakes are systematic, they should be identifiable and susceptible to exploitation by informed investors.  There are a wide range of heuristics that have been suggested, but this study will pay particular attention to \emph{conservatism} and \emph{representativeness}.

@@

Another decision-making short-cut that has been identified by a number of psychological experiments is \emph{representativeness}.  This is the tendency to estimate the probability of an event by reference to its similarity to the parent population or similarity to the process that generated the parent population.  Base effects tend to be downgraded or ignored.  Kahneman presents the example of an accident that takes place in a town that has two taxi companies: 85\% of the cabs are green and 15\% are blue.  A witness identifies the taxi involved in the accident as being blue and tests of the witness reveal that she is correct in her assertion 80\% of the time.  What is the probability that the accident was caused by a blue cab?  Most people put the probability in the range of 80\% to 50\% but it is actually 41\%.

By Bayes' Rule (see Section \ref{secref:ex} for derivation). 
\begin{equation}
P(B|I) = \frac{P(I|B)P(B)}{P(I)}
\end{equation}

>>>>

\documentclass[12pt, a4paper, oneside]{article} % Paper size, default font size and one-sided paper
%\usepackage[dcucite]{harvard}
\usepackage{tikz}
\usetikzlibrary{shapes, shadows, arrows}
\usepackage{rotating}
\usepackage{amsmath}
%\usepackage{setspace}
\usepackage{pdflscape}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage[comma, sort&compress]{natbib}% Use the natbib reference package - read up on this to edit the reference style; if you want text (e.g. Smith et al., 2012) for the in-text references (instead of numbers), remove 'numbers' 
\usepackage{graphicx}
\graphicspath{{./Figures/}} % Specifies the directory where pictures are stored
%\bibliographystyle{plainnat}
\usepackage{listings}
\bibliographystyle{agsm}
\usepackage[colorlinks = true, citecolor = blue, linkcolor = blue]{hyperref}
%\hypersetup{urlcolor=blue, colorlinks=true} % Colors hyperlinks in blue - change to black if annoying
\begin{document}
%\renewcommand[\harvardurl]{URL: \url}
\title{Speculation in foreign exchange: noise or information}
\author{Rob Hayward\footnote{University of Brighton Business School, Lewes Road, Brighton, BN2 4AT; Telephone 01273 642586.  rh49@brighton.ac.uk.  We are very grateful for the comments and suggestions made.  In particular, the EACES conference in Budapest between September 5th and 6th 2014, Andros Steinhoffer and.....}} 
\date{\today}
\maketitle
\begin{abstract}
An event study of extreme speculation is used to test the informational content of speculation.  Speculative extremes are ranked by weight of activity and intensity of sentiment.  In either case, the extreme says little about the future, suggesting that speculation is more than just noise and that there is some fundamental informational content. A unique dataset of risk-reversal skew on option prices is used to measure the weight of speculative sentiment.  This is suplemented with regulatory data on the futures position of speculative traders in the fugtures market.  
\end{abstract}

@@

One of the many things that divided John Maynard Keynes and Milton Friedman was their opinion of speculation.   Keynes main discussion of speculation was in Chapter 13 of \emph{The General Theory} where he identified speuclation as a myopic, sentiment-driven activity that was driven by the desire to ``beat the gun'' or  ``outwit the crowd'' \citep[p. 101]{Keynes1936}.  Friedman, however, say speculation as a stabilising force that would ensure that markets swiftly found equilibrium. In arguing in favour of flexible exchange rates, Friedman asserts that speculators have a stabilsing influence as their success and domination will depend on correctly buying when prices are below fundamental value and selling when above the correct rate.

In Keynes' view, speculative activity will tend to drive prices away from fundamental value as price changes and sentiment feed off each other.  At its most extreme, speculation is likely to cause a series of bubbles.  However, the alternative view regards speculation as an activity that is driven by fundamental knowledge and information and acts to limit overshooting and divergence from fundamental value, thereby reducing volatility.  This second view says that it is the absence of speculators and the liquidity that they provide that encourages excess price movement and volatility.

The aim of this paper is to use the foreign exchange market to assess the relationship between intensity or weight of speculation and the movement of prices.  If speculators are uninformed, their most extreme sentiment and their greatest weight in market activity should tend to cause overshooting and presage bubble-bursting, reversals and a crash; if speculators are informed, activity will drive prices towards fundamental value and equilibrium where volatility and crash risk should be reduced.

From "How noise trading..." in the other folder. 
O'Hara and Saar suggest that there is a difference between noise traders and liquidity traders.  Noise traders exert positive effects on liquidity but weaken price reversals. A tax on trading will reduce noise traders but will also reduce the influence of fundamental traders.

Literature.  Black (1986) highlights the positive effect of noise traders. Shliefer, Summers, DeLong etc identify the limits to arbitrage. The other writings are Battalio, Hatch and Jennings (1997), Harris and Schultz (1998) and Scheinkman and Xiong (2003).  Lowenstein and Willard (2006).  There is considerable debate about the precise role of noise trading in financial markets. There are two strands to the literature on noise trading that emerged in the 1980s:  microstricture literature will use the term "noise traders" and "liquidity traders" interchangeably to identify traders without information.  The motivation of these traders is not identified. Examples of this are Glosten and Milgrom (1985) and Kyle (1985). It is usually assumed to be hedging or liquidity. There is also a \emph{limits-to-arbitrage} literature that uses the term to identify traders that are not movivated by standard explanations of fundamental or liquidity provision. This paper seeks to distinguish between those that trade for liquidty or heging and those that trade because of psychological biases. Fisher Black ``People who trade on noise 
are willing to trade even though from an objective point of view they would be better off 
not trading. Perhaps they think the noise they are trading on is information. Or perhaps 
they just like to trade``.

In theory, traders may trade on mistaken fundamental information.  Shliefer and Summmer (1990) use momentum.  However, Kaniel, Saar and Titman (2006) find evidence of contrarian strategies. This is associated with short-term speculation. Does the activity drive the price away from fundamental value? Stiglitz (1990) is one of the arguments in favour of a Tobin tax.

There is a quesiton about whether experimental markets (like this paper) are superior to actual markets where the identification of trader types is difficult.  Theory - Delong, summers, shleifer etc,  Scheinkman and Xiong (2003).  Lowenstein and Willard (2006).  Empirical work:  (see, for example, Garvey and Murphy [2001], Linnainmaa [2003], Barber et. al. [2004], and Kaniel, Saar, and Titman [2006].

For related papers on securities transaction taxes see Stiglitz [1989], Summers and Summers [1989], 
Amihud and Mendelson [1993], Schwert and Seguin [1993], Umlauf [1993], Subrahmanyam [1998], Dow 
and Rahi [2000], Habermeir and Kirilenko [2003], and Hinman [2003].

Interesting difference between supplying and demanding liqudiity.  Limit orders supply liquidity, market orders demand liqudity.  The activity of liquidity traders will tend to change through the course of the session, moving from supplying to demanding liquidity. Noise traders tend to act in the opposite manner.  They tend to supply more liquidity as the session progresses. Black (1986) noise traders put noise into prices.  The effect of noise traders on liquidity: price impact of trades is lower when noise traders are present.  However, this is mostly in the temporary impact on prices rather than the permanent impact. Therefore, noise traders tend to reduce the amount of volatility in prices because market orders do not temporarily have an effect and then reverse.  The presence of noise traders will increase pricing errors. However, this depends on fundametal traders have valuable information.

There are a number of different ways that noise traders could behave:  activing as SOES bandits (using the NASDAQ small order execution system to make profits) or by using information about the order book; by providing liqudity like a market-maker as a service to liqudiity-traders as in Grossman Miller (1988);  as irrational noise traders, using trading rules or popular models (Shiller 1984, 1990) positive feedback or contrarian.

Footnote -  There is an empirical evidence that individual investors in various countries trade in a contrarian fashion (e.g., Choe, Kho, and Stulz [1999], Grinblatt and Keloharju [2000, 2001], Jackson [2003], Richards [2005], Kaniel, Saar, Titman [2006]). For experimental evidence consistent with contrarian behavior see 
Bloomfield, Tayler, and Zhou (2997)

Can noise traders be viewed as just random noise.  Dow and Gorton (2006) suggest that this is the case and that this is an outcrop of the rational expectations revolution. Informed traders profit by exploiting the diviations from equilibrium. This is the beneifts of Grossman-Stiglitz. What is the source of this noise?  What are the insitiutions? The importance of noise traders here is that they represent a zero-mean, normally distributed random vaiable.  This is also the description that is provided in Kyle (1985).  Liquidity traders trade due to imbalance in the timing of consumption or from portfolio considerations. Shleifer, Summers and DeLong have trading activity that is affected by sentiment, models or "fads".  This is important.  How the average may not be zero. There is a range of behavioural biases that may account for this (see Barberis and Thaler [2003]).The literaure on hetrogenous traders also focuses on noise traders as trend chasers.  These models (see 
for example Brock and Hommes [1998]; Lux [1995]; Lux and Marchesi [2000]) investigate bubbles and 
crashes, and feature fundamental traders versus noise traders. In these models, it is the noise traders who 
induce the aberrant market behavior.

\href{http://papers.ssrn.com.ezproxy.liv.ac.uk/sol3/papers.cfm?abstract_id=869827}{Do noise traders move markets?}  There is strong evidence of herding by individual investors. THere is also momentum. There is reversal:  stocks that were heavily bought one year underperform; stocks that were heavily sold, outperform. The opposite in the short run:  heavily bought one week generate large returns the next week.

\href{http://restud.oxfordjournals.org.ezproxy.liv.ac.uk/content/60/1/1.full.pdf+html}{Campbell and Kyle, Smart money, noise trading and stock market behaviour}. In response to the evidence that the volatilty of stock prices is too great to be a function of discounted expected future earnings (Shiller (1981),  LeRoy and Porter (1981), this paper assesses whether stocks are influnced by exogenous, serially correlated noise. In this model, the noise arises from the interaction of noise traders. 
% Does this now need something on the reason why the foreign exchange market is used?  Could be the Keynes speuclative activity, Fridman paper, liquidity etc.

\subsection{Speculation and noise}
Noise-trading is the label attached to that trading or investment where decisions seem to have been made with no reference to fundamental information; activity is like a random noise.    The DeLong-Summers model of noise trading is calibrated against daily and weekly returns for major foreign exchange rates \citep{Delong1990noise}.  Little support for the hypothesis that extreme levels of sentiment or extreme levels of speculative activity convey information about the future direction of prices is found.  This is something of a contrast to the evidence that has been found in some other markets.  See  \citep{DeBondtOver}, \citep{Chopra1992} and \citep{SteinOptions} amongst others (discussed more fully in Section \ref{secref:empiric} below).

The results of the experiment carried out here suggest that it is not sufficient just to measure the intensity of speculative sentiment or the weight of speculators to determine the divergence from intrinsic value and that the relationship between speculation and price is rather complex.  There is evidence here against the idea of speculation as purely noise.  Rather, it appears that speculation is part of the price discovery process.

The difference in the results achieved here compared to that recorded in some other markets may be due to institutional features. For example, there is a natural tendency for final investors to hold long positions in equities while the foreign exchange market is much more evenly balanced with selling equivalent to buying when the ratio is reversed. The foreign exchange market is also more liquid than the equity market.  The latest survey of activity from the Bank for International Settlements suggests that average daily activity is in the region of \$4.0trn \citet{BISFX}.

\section{Literature review}  
There has been an accumulation of evidence to suggest that financial asset prices do not evolve as a random walk.  Serial correlation or trends are evident suggesting that, contrary to the efficient market hypthesis, future price changes could be partially foreseen by the movement of past price changes.  There are a number of studies that find evidence that the reaction of stocks to information is less than instantaneous.  Ball and Brown seems to be the first study that identifies the upward drift in cumulative abnormal returns after some positive earnings news and downward drift in cumulative abnormal earnings after negative earnings news \citep{BallBrown}.  Bernard provides a summary of the evidence that stocks react slowly to earnings announcements.  He cites twenty examples of ``post earnings announcement drift'' \citep[p. 303]{BernardDrift}.  Part of this under-reaction comes from simply failing to identify what current earnings mean for subsequent quarters.    Bernard argues that the evidence is firmly on the side of earning reactions being too small and taking on average six months to be fully digested \citep[p. 305]{BernardDrift}.

@@

@@

Paul Cootner had already found that extreme returns were more likely to be recorded at the daily frequency than at the monthly level.  This, he suggested was related to the increased ability of speculators to take action at the lower frequency to limit extreme price movement \citep{Cootner1962}. However, the mathematician Benoit Mandelbrot rejected this idea, suggesting that the discrepancy between the evidence from the different frequencies was due to sample size (there being much more high-frequency data).  It was not what he had found with cotton prices.  Mandelbrot had made his own investigation of cotton prices and suggested that serial correlation in the returns was a significant problem and that an alternative distribution would be more appropriate for the returns as there were a number of extreme readings \citep{Mandelbrot1963} and \citep{Mandelbrot1971}.

@@

% Mandelbrot uses what he calls a \emph{stable Paretian} distribution to model returns.  This is a version of Levy's 1925 model and is the same as that used in chapter 4 by ref.  GED or something?  It is a general, stable distribution function with the parameters alpha, beta, x ? representing the tails, skewness, location and wideth.  A beta below 2 suggests flat tails in the distrubution. This is the distrubiton that Mandelbrot and Fama proposed.

\subsubsection{Behavioural finance}

Most relevant for this study are the \emph{heuristics} or short-cuts to decision-making that have been identified by Kahneman, Tversky and other psychologists.   A range of investigations reveal that people deal with the problem of limited information and limited cognitive processing power by using these heuristics  that will economise on time and effort.  The heuristics allow decision to be made quickly with a limited amount of information.  However, they may also create some systematic biases in decision-making that can ensure that, in some situations, investors will make systematic mistakes in their evaluations.  If the mistakes are systematic, they should be identifiable and susceptible to exploitation by informed investors.  There are a wide range of heuristics that have been suggested, but this study will pay particular attention to \emph{conservatism} and \emph{representativeness}.

@@
